---
title: "Exploratory Data Analysis for Health Professionals"
author: Sergio Uribe
output: 
  html_notebook: 
    toc: yes
    toc_float: true
    fig_caption: true
---

# Introduction

# Packages
```{r packages}
library(tidyverse)
```

Data from NHANES
```{r NHANES data}
install.packages("NHANES")
library(NHANES)
data(NHANES)
```

This dataset contains the Body Shape and related measurements from the US National Health
and Nutrition Examination Survey (NHANES, 1999-2004). See
http://www.cdc.gov/nchs/nhanes.htm for details.



# EDA

```{r dim}
dim(NHANES)
```

```{r colnames}
colnames(NHANES)
```


```{r head }
head(NHANES)
```


```{r view dataset}
glimpse(NHANES)
```


```{r summary}
summary(NHANES)
```

### Skim package for EDA
```{r skim package}
install.packages("skimr")
library(skimr)
```

```{r skim all variables}
skim(NHANES)
```


```{r skim some variables}
skim(NHANES, Gender, Depressed, Diabetes, Height, Weight)
```

```{r skim few variables}
NHANES %>% 
         select(Gender, Weight) %>% 
        group_by(Gender) %>% 
        skim()
```



## visdat package for NA
```{r visdat package}
install.packages("visdat")
library(visdat)
```

```{r visdat}
vis_dat(NHANES)
```


## Simple tables
```{r table one factor}
table(NHANES$Gender)
```

```{r table one factor another example}
table(NHANES$Education)
```


2x2 tables

```{r table two factors}
table(NHANES$Gender, NHANES$Diabetes)
```

Add margin totals

```{r table with margins totals}
addmargins(table(NHANES$Gender, NHANES$Diabetes))
```

Proportion table

```{r prop table}
prop.table(table(NHANES$Gender, NHANES$Diabetes))
```

As percentage
```{r prop table as %}
prop.table(table(NHANES$Gender, NHANES$Diabetes))*100
```

with two decimals

```{r limit decimals}
options(digits = 3)
prop.table(table(NHANES$Gender, NHANES$Diabetes))*100
```

Combining addmargins and prop.table

```{r table combined}
addmargins(prop.table(table(NHANES$Gender, NHANES$Diabetes))*100)
```

```{r n x n x n tables}
table(NHANES$Gender,  # rows
      NHANES$Education,  # columns
      NHANES$Diabetes) # factor
```
#### Other tables packages
With xtabs
```{r xtabs}
xtabs(~ Gender +
              Education +
              Diabetes,
      data=NHANES)
```

With ftables
```{r ftable}
ftable(NHANES$Gender,  # rows
      NHANES$Education,  # columns
      NHANES$Diabetes) # factor
```




#### Nice Output tables
```{r gmodels}
install.packages("gmodels")
library(gmodels)
CrossTable(NHANES$Gender, NHANES$Diabetes, 
           digits = 2, 
           prop.r = T, # If TRUE, row proportions will be included
           prop.c = T, # If TRUE, column proportions will be included
           prop.t = F, # If TRUE, table proportions will be included
           prop.chisq = FALSE, # If TRUE, chi-square contribution of each cell will be included
           chisq = TRUE, #If TRUE, the results of a chi-square test will be included
           dnn = c("Gender", "Diabetes"))
```


## Visualizations

### Continuous data
```{r histogram}
NHANES %>% 
        ggplot(aes(x = Age)) +
        geom_histogram()
```

Change the bins

```{r change bins to 5}
NHANES %>% 
        ggplot(aes(x = Age)) +
        geom_histogram(bins = 5)
```

```{r change bins to 10}
NHANES %>% 
        ggplot(aes(x = Age)) +
        geom_histogram(bins = 10)
```

```{r scatterplot}
NHANES %>% 
        ggplot(aes(x = Height, y = Weight)) +
        geom_point()
```

```{r add regression line}
NHANES %>% 
        ggplot(aes(x = Height, y = Weight)) +
        geom_point() +
        geom_smooth()
```


```{r add regression line}
NHANES %>% 
        filter(!is.na(DirectChol)) %>% 
        ggplot(aes(x = Height, y = Weight, color = Gender)) +
        geom_point(alpha = 0.3)
```



### Categorical data

#### Barplots
```{r barplots}
NHANES %>% 
        ggplot(aes(x = Diabetes)) +
        geom_bar()
```

```{r order axis}
NHANES %>% 
        ggplot(aes(x = Education)) + 
        geom_bar()
```
Flip categories

```{r flip axis}
NHANES %>% 
        ggplot(aes(x = Education)) + 
        geom_bar() + 
        coord_flip()
```

Eliminate NAs

```{r ploy without NAs}
NHANES %>%
        filter(!is.na(Education)) %>% 
        ggplot(aes(x = Education)) + 
        geom_bar() + 
        coord_flip()
```

Add a factor
```{r ploy without NAs}
NHANES %>%
        filter(!is.na(Education)) %>% 
        ggplot(aes(x = Education, fill = Gender)) + 
        geom_bar() + 
        coord_flip() 
```

Nicely formated
```{r ploy without NAs}
NHANES %>%
        filter(!is.na(Education)) %>% 
        ggplot(aes(x = Education, fill = Gender)) + 
        geom_bar() + 
        coord_flip() +
        theme_minimal() + 
        labs(title = "Educational level by gender", 
             subtitle = "Data fron NHANES, 2014", 
             y = "Count", 
             x = "Educational level")
        
```

#### By groups
```{r one group}
NHANES %>% 
        ggplot(aes(x = Height)) +
        geom_histogram()
```


```{r two groups}
NHANES %>% 
        ggplot(aes(x = Height, fill = Gender)) +
        geom_histogram()
```

```{r add alpha}
NHANES %>% 
        ggplot(aes(x = Height, fill = Gender)) +
        geom_histogram( alpha=.7, 
                        bins = 15) 
```
#### Boxplots

```{r boxplots}
NHANES %>% 
        ggplot(aes(x = Gender, y = Height)) +
        geom_boxplot()
```

Dirty option: limit the coordinates
```{r}
NHANES %>% 
        ggplot(aes(x = Gender, y = Height)) +
        geom_boxplot() + 
        scale_y_continuous(limits=c(75,150))
```

More elegant solution, with R Base

```{r}
boxplot(NHANES$Height~NHANES$Gender,
        horizontal = F,
        axes = T,
        outline = FALSE)
```


#### Categorical data independence

```{r}
mosaicplot(table(NHANES$Gender, NHANES$Diabetes))
```

```{r}
mosaicplot(table(NHANES$Gender, NHANES$Diabetes), shade = T)
```

Exploring some associations

```{r}
mosaicplot(table(NHANES$Diabetes, NHANES$PhysActive), shade = T)
```


## Numerical summaries
For all variables
```{r}
summary(NHANES)
```

For one variable
```{r}
summary(NHANES$Age)
```

### Create table summaries

```{r}
NHANES %>% 
        group_by(Gender) %>% 
        summarise(n = n(), "Mean age" = mean(Age), SD = sd(Age))
```

